---
title: Brexit
author: JB
date: '2018-12-15'
slug: brexit
categories:
  - Brexit
  - Intelligence Analysis
  - decision-making
tags: []
---

```{r setup, include=FALSE}
library(kableExtra)
library(knitr)
x<-data.frame(
  Reamin = c("Continued collaboration with EU nations", "Open borders that allow access to more skilled workers", "Simple goods transfer within a single customs union","Greater negotiating power with external nations"),
  Leave   = c("More control over immigration","Saving from EU transfer payments","Greater control over trade negation","More flexibility in policy creation")
)
```
Back again! And what a month we have had in the UK. In case you didn’t know, I used to work in Canadian Politics, which compared to the rest of the world (especially these days) is quite tame. But politics is something I am passionate about. I like to think of myself as civilly minded democrat committed to reflecting the will of the people into robust and positive policy. Others might disagree, you know who you are ;). But I think we can all agree that Brexit has failed to meet expectations and is what I want to focus on today.

This post is not about politics directly, nor do I talk about my personal stance on Brexit. Rather, this is a case study in what not to do in decision making and the consequences that result. Let’s start back in June 2016. I stepped into a voting booth in Greenhithe, Kent (yes, Canadians can vote in the UK without being citizens!), and I was greeted with this question:

#### Should the United Kingdom remain a member of the European Union or leave the European Union?

##### &#9634; Remain a member of the European Union

##### &#9634; Leave the European Union

Pretty straightforward right? Should I stay or should I go? But what are the criteria for that decision? Imagine a simple pros/cons list example. I don’t care if you don’t like my examples, they are purely illustrative. 

```{r, echo = FALSE}
kable(x)
```

So that is my list, and we might surmise that my primary interests are immigration, trade and policy. But now imagine that 66 million other people are also making this list and boiling that criteria down. You can imagine with that many lists there will be more than three items.

However, notice that there is an asymmetry of information here. Remain is the default position so I know what I am getting. I can measure and evaluate those items. And even if I wanted reform, remain is a base to start with data to compare. Now, look at the leave-side. “More control over immigration.” What does that look like? How do I define “control”? How do 66 million people define it? Are they all the same? I think not. This is a failure to articulate a hypothesis about the future or even communicate a desired state, let alone synchronise that across process across an entire population.

Fast forward to December 2018 and the anarchy that the Brexit proposal as wrought. I don’t think that it is controversial to say as both “remainers” and “leavers” are disenfranchised. Why is that? Because the question they are asked was too simplistic for the size and complexity of the decision to be made.

I want to give a shout out to the excellent work of the UK electoral commission here. They did a fantastic job considering that they needed to boil down a multi-national policy to one simple sentence that anyone could both read and interact with. They even wrote a [56-page report](https://www.electoralcommission.org.uk/find-information-by-subject/elections-and-referendums/past-elections-and-referendums/eu-referendum/eu-referendum-question-assessment) emphasising the need for both neutral and clear wording and warned of the dire consequences of communicating something so complex in a binary format. They did a great job and should be commended.

However, we still need to live with the fallout, and here we are. “Leavers” feel betrayed by the deal because it doesn’t go far enough in protecting the UK and severing ties from the EU. “Remainers” never wanted this in the first place. How can everyone not be happy? One side won, didn’t they? But no one ever said, “how should we leave” they just said, “we should leave because of x,y,z.” That is not a solution or forecast we can judge, but a statement. While a statement might be based on data, it doesn’t speak to what an alternative might look like or what consequences might result from an alternative, which is critical when making a comparative judgement. One might argue that that is for politicians to decide, but you have just handed them an impossible task. Here is an example from Dan Ariely’s book, Predictably Irrational. By SEC decree, US CEO’s were required to publically disclose their salaries to drive transparency and encourage moderation. But, CEO’s now knew what their counterparts were making, encouraging them to demand higher pay since now they had hard data telling them they were underpaid. CEO salaries increased approximately three-fold between 1976 and 1993, going from 36 times the average worker pay to 131 times the average worker pay.

Imagine a politician taking those 66 million pro/con lists and synthesizing them in one coherent, commonly accepted Brexit policy. Even if you cater to the majority there will be concessions because there is an omnipresent disconnect between preference and consequence. Everything has consequences, good or bad because as something changes so does context. You buy a dog; you have a friend, but you need to pick up its poo. You buy a car; you can travel farther, but you need to keep it repaired and fuelled. You leave the EU to have greater control of your domestic affairs; there could be market volatility, medicine shortages, work shortages, etc. The same can be said for preference and tolerance. One person might be fine with high food prices if they have greater say over immigration, while another may only accept high food prices if they can guaranty that money goes to benefits programmes for citizens. 

The point I am aiming for is that the UK people were asked to make a very large and complex decision on the basis of a binary choice, with few facts and a lot of theorizing and preference taking and now… no one sees the outcome they had envisioned. The same thing happens in intelligence. The decision maker has big ideas, a vision of the end product and the support they need, and the analyst misses the mark entirely, creating a divide between needs and delivery. The “It’s great but not what I asked for” problem from previous posts. 

I think Brexit is an excellent analogy for decision dissonance where requirements are not captured, analysis not conducted and expectations not met. Regardless of your position on the political spectrum or the Brexit debate, we need to accept that we did not get off to a great start as well as acknowledging a current divide, not just between political ideas, but perceptions of the best way forward based on solid evidence. Issues like the Irish Backstop highlight this problem. The official policy mechanisms ‘could’ lock the UK into the EU customs union indefinitely, or a solution might be reached in a few years’ time and this is an appropriate compromise to reach a leaving framework. But the lack of certainty is the problem. Had voters knew about this two years ago would they vote the same? Ask the same questions? At the end of the day, everyone in the UK will be affected by the mammoth policy regardless of ideology and we owe it to fellow citizens and residents to make the best choices possible to ensure our personal liberty and opportunity are protected.

For us intelligence folk, remember that the client wants simple solutions but needs to accept there are complex answers and knotty problems. Failure to communicate that complexity, when needed, can result in failed intelligence that actually detracts from positive choices and creates miss information or unrealistic expectations. It might be cheeky, but next time you’re faced with a communicating a difficult choice to a decision maker think, “How could this go the way of Brexit?”

